Caring 15-minute neighbourhoods

### 2024-10-10 paper/Manuscript.qmd
Draft to submit to journal. And some repo clean-up.

---
title: "Identifying 15-minute neighbourhoods with complete caring potential"
author:
  - name: Anastasia Soukhov 
    email: soukhoa@mcmaster.ca
    affiliations:
        - name: McMaster University
          department: School of Earth, Environment & Society
          city: Hamilton, Canada
    attributes:
        corresponding: true
  - name: Léa Ravensbergen
    email: ravensbl@mcmaster.ca
    affiliations:
        - name: McMaster University
          department: School of Earth, Environment & Society
          city: Hamilton, Canada
  - name: Lucía Mejía Dorantes
    email: lucia.mejia.dorantes@gmail.com
    affiliations:
        - name: Consultant
          city: Karlsruhe, Germany
  - name: Antonio Páez
    email: paezha@mcmaster.ca
    affiliations:
        - name: McMaster University
          department: School of Earth, Environment & Society
          city: Hamilton, Canada
format: 
  docx
  # elsevier-pdf:
  #   keep-tex: true
  #   include-in-header:
  #     text: |
  #       \usepackage{graphicx}
  #       \usepackage{unicode-math}
  #       \usepackage{hyperref}
  #       \def\tightlist{}
  #       \usepackage{setspace}
  #       \doublespacing
  #       \usepackage{lineno}
  #       \linenumbers
  #   journal:
  #     name: 7th International Conference on Women and Gender in Transportation (WGiT)
  #     formatting: preprint
  #     model: 3p
  #     cite-style: authoryear
keywords: ["Accessibility", "Mobility of Care", "Gender", "15-minute neighbourhood", "Care Complete Neighbourhoods"]
bibliography: bibliography.bib
---

@@

summary_df$name <- c("GROCERY TOTALS", "Convenience Store", "Grocer",
                     "DEPENDENT TOTALS", "Community or Rec. centre", "Daycare or EarlyON","LTC or retirment home","Park","School","Senior centre",
                     "HEALTH TOTALS","Dentist","Hospital or clinic","Pharmacy",
                     "ERRAND TOTALS","Bank or ATM","Library", "Post office",
                     "Category diversity")

@@

The "15-Minute Neighbourhood" concept has been adopted by leaders around the world as a way to promote human-scale cities. The concept emphasizes a transport and land-use mix that supports short-trips to daily necessities by foot or bicycle from ones' residence. This study adopts the 15-minute neighbourhood concept through the lens of "Mobility of Care", a framework that foregrounds travel to care destinations, travel done predominately by women, in contrast to the better-studied travel to employment and leisure. While the 15-minute neighbourhood concept is flexible enough to consider all destination types, gendered examinations are relatively lacking in the literature and no research to date has focused explicitly on care destinations. In this study, we address these gaps by identifying which areas in Hamilton, Canada have the potential to be ‘caring 15-minute neighbourhoods'. To do so, a database of care destinations was compiled and used to estimate both the number (cumulative opportunities indicator) and diversity of mix (entropy measure) of destinations reachable within a 15-minute walk from all residential parcels. Parcel-level results were classified into 'caring' (quantity of reachable opportunities) and 'complete' (diversity of care destinations) neighbourhood typologies of varying degrees using machine learning techniques (Self-Organizing Maps (SOM) and hierarchical clustering). Inequalities in the relationship between the identified caring typologies and social-economic indicators of who currently reside in the neighbourhoods were then explored. Our results suggest the majority of 'caring and complete' neighbourhoods are within the urban core where a higher proportion of lower-income households reside. Contrawise, neighbourhoods that are 'not caring and not complete' are located in peripheral regions that tend to have higher household incomes. Further, certain neighbourhoods in suburban and rural communities are 'caring and somewhat complete', showing the potential to become 'complete' through possible urban policy intervention. Implications of neighbourhoods that have a high potential to be improved, and socio-demographic characteristics associated with income such as single-parent households and economic poverty are discussed. Taken together, this study offers a theoretical bridge to connect Mobility of Care and the 15-minute neighbourhoods concept, a data-driven classification and clustering methodology to assess neighbourhoods on a 'caring and complete' continuum, and discussion of implications relevant to gender mainstreaming within transportation planning.

@@

Though the term 15-Minute City is relatively new, neighbourhood planning practice aiming to foster community and local travel to amenities is not. As recent examples, planned neighbourhoods dominated post-WWII built urban form: the Neighbourhood Unit Concept in the western world [@brodyNeighbourhoodUnitConcept2013] and a parallel concept of the “mikrorayons” neighbourhoods in the Soviet Union  [@kissfazekasCircleParadigms15minute2022]. Along different dimensions, these planning forms have been extensively critiqued [@talenSocialSciencePlanned2017]. In this respect, neighbourhood planning approaches such as the 15-Minute City concept are being re-embraced by some, but from bottom-up perspective using co-creation tools and meaningful resident participation to achieve equitable and justice neighbourhood forms [@mahmoudCocreationPathwayUrban2021]. How to equitably prescribe urban form that 'enables' contact with social opportunities (instead of 'engendering it' from the top-down) is still under debate. For instance, questions such as _what_ are destinations that matter, by what mode and travel time threshold, and _for whom_ remain under discussion among proximity-based planners who frequently use accessibility-based tools [@silvaProximitycentredAccessibilityConceptual2023; @silvaRegionalAccessibilityUndermining2022; @guzmanProximityEnoughCritical2024].

From another perspective, feminist geographers, urban planners and other research groups agree that urban structure influences travel, but city planning fails short when planning for a neutral identity. In contrast, cities ought to be planned according to the multiple identities of the inhabitants [@vacchelli_towards_2018; @urbandevelopmentviennaGenderMainstreamingUrban2013]. Researchers have demonstrated how mobility patterns differ according to gender [@lawWomenTransportNew1999; @cresswell_gendered_2008; @levy2013travel; @little_gender_1994; @tronto_toward_1990]. Gender and social norms and cultural factors play a role in how, where, with whom, what for, and how far we move. Women have higher activities related to care [@ilo_care_2018], no matter at what life stage they are [@garciaromanGenderDifferencesTime2022]. For all these reasons the spatial organization of activities has a different impact according to gender, the 15-Minute City concept cannot be gender blind.

To clarify questions surrounding the 15-Minute City concept of what destinations matter and for whom, this work builds a theoretical bridge to "Mobility of Care". The Mobility of Care term was coined in @sanchezdemadariagaMobilityCareIntroducing2013; it refers to all travel required to sustain the needs of a household, such as grocery shopping, escorting children, travelling to health appointments, and running errands.  While decades of research have examined types of household-serving trips (e.g., shopping trips, escorting to school), @sanchezdemadariagaMobilityCareIntroducing2013 was the first to consider them all as one category. In doing so, Mobility of Care is shown to be a significant proportion of daily travel, characterizing the trip purpose of approximately 30% of adults’ trips [@sanchezdemadariagaMobilityCareIntroducing2013; @sanchezdemadariagaMeasuringMobilitiesCare2019; @ravensbergen2023exploratory]. We argue Mobility of Care is especially relevant to the 15-Minute City, as care trips are necessary for _all_ people and they most often occur at the local level. Moreover, care trips take on tremendous gendered and socio-economic importance, as women, and especially those from lower income households, complete the majority of care trips [@ravensbergen2023exploratory]. Despite care travel's importance in advancing gender equality and equity broadly, the examination of mobility of care and access to care destinations have been under-examined relative to employment destinations. This is especially pertinent in accessibility research that has largely focused on employment points of interest e.g., [@farberOntarioLineSocioeconomic2019; @duarteInfluenceJobAccessibility2023; @ryanAccessibilitySpaceTime2023].

The objectives of this research are two-fold. First, we aim to theoretically bridge Mobility of Care with the 15-Minute City concept to to re-imagine what local amenities matter and define an associated accessibility and diversity of accessibility measure. Second, taking a bottom-up data-driven approach, we apply a classification and clustering machine learning methodologies to support the bottom-up identification of _15-minute caring neighbourhoods_ through an empirical examination of the mid-cited city of Hamilton, Canada. These neighbourhoods are then interpreted on the continuum of having (and not having) the region-relative land-use and transportation infrastructure to support _completely caring_ access. Who currently resides in which neighbourhood is examined, and ways to further expand the care accessibility story are discussed.

@@

In the last decades, the need for more sustainable, healthier, and livable cities have received significant attention. Planners and decision-makers have proposed alternatives to mobility-based neighbourhood planning principles to ones that center proximity to urban functions [@pozoukidou15MinuteCityDecomposing2021]. In this context, the 15-Minute City is now in the public spotlight [@logan_xminute_2022; @moreno_introducing_2021]. In @moreno_introducing_2021 where the 15-Minute City concept was introduced, four universal dimensions are outlined: density (in terms of people per km$^2$), diversity (including mixed-land use and diversity of people), the temporal and spatial proximity to essential services, <!-- (residential areas, work, commercial areas, education centres, health facilities and other basic institutions)--> and digitalisation. However, the framework presented has been criticized within academic and planning circles [@guzmanProximityEnoughCritical2024]. Critiques highlight the concept falls short in addressing the  pre-existing structural forces and individual characteristics that drive inequalities, influencing who benefits or could benefit the most from such a paradigm [@dimarino15minuteCityConcept2023; @willberg15minuteCityAll2023]. A 15-Minute City for all people is an aspirational ideal, but how will existing mobility and accessibility inequalities be addressed. Without directed and context-specific solutions, this increasingly popular concept risks becoming a place-branding slogan [@pozoukidou15MinuteCityDecomposing2021].

Reflecting the flexible and aspirational presentation of the 15-Minute City; cities that have adopted the concept have done so using a diverse range of definitions and tools along with indistinct universal approaches. A trail blazer was Portland, US in the Portland Plan [@city_of_portland_20minute_2010] of April 2012. Adopted before the '15-Minute City' concept of today, it aims to foster an inclusive city development based on prosperity, education, health, and equity over a 30 year horizon. The plan aims to promote neighbourhood self-sufficiency and connectivity to city centers and centres of employment. The progress report describes a high-level focus on equitable service delivery to all residents with equity concerning topics related to racial equity and people with disabilities [@portland_government_portland_2017], similarly taking on an "all populations" approach. Later, other cities also incorporated proximity-based goals into their plans, adopting similar neutral entity approaches. For instance, in the plan that later catapulted the 15-Minute City into public discourse: Paris mayor Anne Hidalgo proposed a 15-minute city plan in her 2020 re-election campaign that foregrounded six social functions that should be easily accessible from any location, namely: housing, work, health care, groceries, education and leisure [@ville_de_paris_paris_2022]. These ideas inspired language in the agendas of other cities in the Western world such as Ottawa, in the Canadian context, who also adopted a 15-Minutes approach in their recent Official Plan [@ville_dottawa_quartier_2021]. From the review of worldwide practice literature in @teixeiraClassifying15minuteCities2024, the 15-Minute City concept is in early phases of implementation around the world and the diverse range of definitions, strategies, and instruments present a significant knowledge and implementation gap: the 15-Minute City is aspirational but _how do we get there and whom will benefit_?

@@

The re-defined bottom-up approach to community and neighbourhood planning has been adopted by New Urbanists, from which the 15-Minute City stems [@kissfazekasCircleParadigms15minute2022]; however, the effectiveness of these bottom-up processes remain to be seen due to the contemporary nature of the concept. Though, related research can be examined. In recent years, the question of _how can physical form be planned to enable an improved quality of life, and for whom with what outcomes_ has occupied the urban planning research agenda. For instance, the examination of low-income households access to transportation and their likelihood of gaining employment [@blumenbergDriveWorkRelationship2017; @bastiaanssenDoesBetterJob2022] or the relationship between children's access to public transit and participation in after-school activities [@palmRolePublicTransit2020]. A new wave of researchers and practitioners focused on local and context-specific relationships with the proximity to destinations have also emerged [@silvaProximitycentredAccessibilityConceptual2023; @silvaRegionalAccessibilityUndermining2022]. As reviewed  in the city plans that have adopted 15-Minute City approaches, the NUC's criticisms, and recent urban planning research, the question of _how to enable improved quality of life through urban built environment_ has shown to be highly context-sensitive, with the need for further investigation.

@@

In examining how to enable improved quality of life through urban built environment, an increasingly popular tool have been accessibility measures. These measures are a way to quantify the ease of reaching destinations from a given point in space and have been used to examine urban areas through just and sustainable city planning agendas [@vale2023accessibility; @handyAccessibilityIdeaWhose2020]. As the 15-Minute City is an amenity-provision neighbourhood planning concept which aims to enable urban environment that improves life quality, its analysis lends well to the use of accessibility tools [@guzmanProximityEnoughCritical2024]. Foreseeably, recent works have applied accessibility measures to investigate the 15-Minute City across various geographic scopes. For instance, in Naples, Italy [@gaglioneUrbanAccessibility15minute2022], Barcelona, Spain [@graellsCityCitiesMeasuring2021], Vancouver, Canada [@hosford15minuteCityReach2022], and in urban areas across Europe [@vale2023accessibility]. The "cumulative opportunity" measure has been applied to many 15-Minute City accessibility examinations. This form of accessibility quantifies how many destinations can be reached from a point in space within a given travel time threshold, pairing well with examinations that have a normative travel time threshold, like the 15-Minute City. Furthermore, the cumulative opportunity measure is widely appreciated for its intuitive computation and popularity among practitioners [@handyAccessibilityIdeaWhose2020; @handyMeasuringAccessibilityExploration1997; @chengInvestigatingWalkingAccessibility2019]. However, accessibility measures other than the cumulative opportunity have also been applied, reflecting the diversity of measures in the literature [@guzmanProximityEnoughCritical2024].

Another method which compliments the assessment of the 15-Minute City as it relates to the diversity of opportunity types, is entropy. The entropy measure has been widely used to characterise the diversity of land-use mix in early work by @frankLinkingObjectivelyMeasured2005 based on the species diversity Shannon-Wiener index that expresses relative evenness throughout a sequence [see review in @whittaker1972evolution]. It has also been used in social sciences to comprehend the occurrence of social events in a structured manner specially with regard to life trajectories [@ritschardSequenceAnalysisRelated2018]. In the realm of urban studies and planning literature, the malleable concept of entropy has been used to characterise the diversity in land-use mix [@ewingTravelBuiltEnvironment2010a], especially in the context of active modes [@luUrbanDensityDiversity2017; @mavoaIdentifyingAppropriateLanduse2018] and suburban sprawl [@randallGISBasedLand2015]. As well, it has been used to understand mobility behaviour [@mcbride2020exploration; @montero2023applying; @montero2023role]. Work that has used diversity indices alongside accessibility analysis are scarce, though examples of its use as parameters within accessibility scores in the case of employment opportunities [@chengMeasuringUrbanJob2013; @daiIncorporatingJobDiversity2018] and more recently to describe the diversity in transit facilities [@yinIncorporatingFacilityDiversity2024] are present. To the authors knowledge, the use of an entropy measure to reflect the diversity in care destination amenity types and in context within a 15-minute neighbourhood, has yet to be done. However, entropy measures theoretically compliment the 15-Minute City concept.

In addition, classification algorithms have been useful tools in the identification of mobility and spatial typologies in relation to transportation planning. Machine learning approaches have often been deployed. For instance, @burkeGeospatialAnalysisFramework2022 uses spatially constrained multivariate clustering, a regionalisation technique, to develop urban form typologies related to the 15-Minute City. The k-means algorithm is another popular method; @ganUnderstandingUrbanMobility2020 applies it to analyse travel behaviour and classify metro stations based on mobility patterns. Similarly, [@montero2023role; @montero2023applying] cluster travel behavior profile typologies representing different population segments and place of residence in Barcelona Metropolitan Area. Combing classification techniques is also common; for instance, @delmelleDifferentiatingPathwaysNeighborhood2017 uses self-organizing maps (SOM) algorithm to group US neighbourhoods by minimizing dissimilarity in attributes before applying k-means for further classification. Similarly, @victorianoTimeSpaceMoney2020 uses SOM to classify individuals' travel patterns according to the dissimilarity in mobility attributes and uses the decision tree algorithm to partition the data into interpretable classifications.

All-in-all, within the context of examining the 15-Minute City, a variety of tools have been applied in the literature. However, the question of _15-minute cities for whom_ sill demands further examination. This work aims to bridge the concept of the 15-Minute City with another normative framework: Mobility of Care.

@@

Rather than focusing on _all_ destinations, it may valuable to examine those related to caring activities. This re-framing offers a feminist perspective on  _what_ urban functions matter, connecting well with the 15-Minute City concept. Caring activities, which meet the physical, psychological, and emotional needs of others, are among the most essential and fundamental activities in society [@ilo_care_2018]. Yet, they are the most unequal, undervalued, and even devalued activities worldwide. Conventionally, caring activities have been borne on women’s shoulders [@haydenGrandDomesticRevolution1982; @hochschild_second_2012]. According to @ilo_care_2018, women and girls perform more than three-quarters of the total amount of unpaid care worldwide, a gender gap that varies geographically [@RN5]. This unequal share of caring responsibilities leads to multifaceted gendered differences: in career development, profession selection, contract type, pay gap, and time poverty, as recognized by various international organisations [@EIGE2016; @ilo_care_2018]. In terms of spatial and transportation planning, almost one third of daily trips are for care purposes [@sanchezdemadariagaMobilityCareIntroducing2013; @sanchezdemadariagaMeasuringMobilitiesCare2019; @ravensbergen2023exploratory]. From this research motivation, Sánchez de Madariaga coined the term Mobility of Care in 2013 to refer to all travel required to sustain the needs of a household, such as grocery shopping, escorting children, travelling to health appointments, and running errands [@sanchezdemadariagaMobilityCareIntroducing2013]. While an undercurrent of research had examined these unique household-serving trips over the decades, her seminal work was the first to consider all these trips as one category and demonstrate how mobility of care is a significant proportion of daily travel.

The Mobility of Care concept also explicitly integrates inter-sectional equity considerations. In considering Mobility of Care within the 15-Minute City one could address a common criticism leveled at the 15-Minute City [@guzmanProximityEnoughCritical2024]. Perhaps unsurprisingly given the gendered division of care work discussed, women have consistently been found to complete more mobility of care trips than men. In one study, mobility of care comprised 32% of women’s daily trips compared to 28% of men’s. While this gap is significant, it was found to be far greater in lower income households where women complete 20% more care trips than men [@ravensbergen2023exploratory]. Sánchez de Madariaga not only shows how important these mobility of care trips are, but also highlights the ways in which “mobility of care is systematically under-represented in any analysis of urban transport” (p. 37). As a product of the masculinst bias in planning, transport surveys and tools most of the times do not directly capture mobility of care, which re-enforces the idea that these trips are not a significant part of daily mobility. In this respect, the feminist perspective of the cities of proximity is still underestimated with only few examples on the topic [@gil_sola_choose_2022; @macintyre_her_2022].

@@

Hamilton also exhibits spatial disparities in social and economic indicators; their spatial distribution is visualised in @fig-Fig2. The densely populated inner city is characterised by lower average incomes, and a higher prevalence of households living under the low-income cutoff thresholds (LICO). Noteably, the suburban areas of the city tend to have a greater proportion of children and a lower proportion of one parent households.

>>>>

---
title: "Caring 15-minute neighbourhoods"
author:
  - name: Anastasia Soukhov 
    email: soukhoa@mcmaster.ca
    affiliations:
        - name: McMaster University
          department: School of Earth, Environment & Society
          city: Hamilton, Canada
    attributes:
        corresponding: true
  - name: Léa Ravensbergen
    email: ravensbl@mcmaster.ca
    affiliations:
        - name: McMaster University
          department: School of Earth, Environment & Society
          city: Hamilton, Canada
  - name: Lucía Mejía Dorantes
    email: lucia.mejia.dorantes@gmail.com
    affiliations:
        - name: Consultant
          city: Karlsruhe, Germany
  - name: Antonio Páez
    email: paezha@mcmaster.ca
    affiliations:
        - name: McMaster University
          department: School of Earth, Environment & Society
          city: Hamilton, Canada
format: 
  # docx
  elsevier-pdf:
    keep-tex: true
    include-in-header:
      text: |
        \usepackage{graphicx}
        \usepackage{unicode-math}
        \usepackage{hyperref}
        \def\tightlist{}
        \usepackage{setspace}
        \doublespacing
        \usepackage{lineno}
        \linenumbers
    journal:
      #name: "Transportation Research Part A: Policy and Practice"
      formatting: preprint
      model: 3p
      cite-style: authoryear
keywords: ["Accessibility", "Mobility of Care", "Gender", "15-minute neighbourhood", "Care Complete Neighbourhoods"]
bibliography: bibliography.bib
---

@@

summary_df$name <- c("GROCERY TOTALS", "Convenience Store", "Grocer",
                     "DEP. TOTALS", "Comm. or Rec. centre", "Daycare or EarlyON","LTC or retirment home","Park","School","Senior centre",
                     "HEALTH TOTALS","Dentist","Hospital or clinic","Pharmacy",
                     "ERRAND TOTALS","Bank or ATM","Library", "Post office",
                     "Category diversity")

@@

The "15-Minute City" concept has been embraced by global leaders to promote human-scale neighbourhoods with transport and land-use designs that support short trips to daily necessities. This paper bridges the 15-Minute City to "Mobility of Care",  a framework that foregrounds travel to care destinations, travel done predominately by women. This focus contrasts the more commonly studied travel to employment and leisure destinations. While the 15-Minute City concept is flexible enough to consider all destination types, gendered examinations are relatively lacking in the literature, and little research to date has focused explicitly on care destinations. This gap is addressed in this paper by identifying which areas in the city of Hamilton, Canada are ‘caring 15-minute neighbourhoods’. To do so, a database of care destinations was compiled to estimate the number (using the cumulative opportunities accessibility measure) and diversity of mix (using the entropy measure) of care destinations within a 15-minute walk from residential parcels. Using data-driven machine learning techniques (Self-Organizing Maps and Decision Trees), neighborhoods were classified into 'caring 15-minute neighbourhood' typologies that are examined across residential socio-economic profiles. Our results suggest that the majority of caring 15-minute neighbourhoods are in the urban core, where more lower-income households reside. In contrast, not caring 15-minute neighbourhoods in higher-income peripheral areas. Areas that make good candidates for urban policy intervention are identified, and the implications of enhancing 15-minute walkable caring access are discussed in relation to gender mainstreaming in transportation planning.

@@

Though the term 15-Minute City is relatively new, neighbourhood planning practice aiming to foster community and local travel to amenities is not. As recent examples, planned neighbourhoods dominated post-WWII built urban form: the Neighbourhood Unit Concept in the western world [@brodyNeighbourhoodUnitConcept2013] and a parallel concept of the “mikrorayons” neighbourhoods in the Soviet Union  [@kissfazekasCircleParadigms15minute2022]. Along different dimensions, these planning forms have been extensively critiqued [@talenSocialSciencePlanned2017]. In this respect, neighbourhood planning approaches such as the 15-Minute City concept are being re-embraced by some, but from bottom-up perspective using co-creation tools and meaningful resident participation to achieve equitable and justice neighbourhood forms [@mahmoudCocreationPathwayUrban2021]. How to equitably prescribe urban form that 'enables' contact with social opportunities (instead of 'engendering it' from the top-down) is still under debate. Questions remain, including: what destinations really matter?, what mode and travel time threshold is most appropriate?, and who benefits? For instance, these dimensions are under discussion among proximity-based planners who use accessibility-based tools [@silvaProximitycentredAccessibilityConceptual2023; @silvaRegionalAccessibilityUndermining2022; @guzmanProximityEnoughCritical2024].

From another perspective, feminist geographers, urban planners and other researchers agree that urban structure influences travel, but city planning falls short in planning for a neutral identity. In contrast, cities ought to be planned according to the multiple identities of the inhabitants [@vacchelli_towards_2018; @urbandevelopmentviennaGenderMainstreamingUrban2013]. Researchers have demonstrated how mobility patterns differ according to gender [@lawWomenTransportNew1999; @cresswell_gendered_2008; @levy2013travel; @little_gender_1994; @tronto_toward_1990]. Gendered social norms and cultural factors play a role in how, where, with whom, what for, and how far we move. Women have higher activities related to care [@ilo_care_2018], no matter at what life stage they are [@garciaromanGenderDifferencesTime2022]. For all these reasons the spatial organization of activities has a different impact according to gender, the 15-Minute City concept should not be gender blind.

This work builds a theoretical bridge between the 15-minute city and mobility of care concepts to answer questions about what destinations matter and who benefits from proximity planning. The Mobility of Care term was coined in @sanchezdemadariagaMobilityCareIntroducing2013; it refers to all travel required to sustain the needs of a household, such as grocery shopping, escorting children, travelling to health appointments, and running errands.  While decades of research have examined types of household-serving trips (e.g., shopping trips, escorting to school), @sanchezdemadariagaMobilityCareIntroducing2013 was the first to consider them all as one category. In doing so, Mobility of Care is shown to be a significant proportion of daily travel, characterizing the trip purpose of approximately 30% of adults’ trips [@sanchezdemadariagaMobilityCareIntroducing2013; @sanchezdemadariagaMeasuringMobilitiesCare2019; @ravensbergen2023exploratory]. We argue Mobility of Care is especially relevant to the 15-Minute City, as care trips are necessary for _all_ people and they most often occur at the local level. Moreover, care trips take on tremendous gendered and socio-economic importance, as women, and especially those from lower income households, complete the majority of care trips [@ravensbergen2023exploratory]. Despite care travel's importance in advancing gender equality -and equity more broadly-, mobility of care and access to care destinations have been under-examined relative to employment destinations. This is especially pertinent in accessibility research that has largely focused on employment points of interest e.g., [@farberOntarioLineSocioeconomic2019; @duarteInfluenceJobAccessibility2023; @ryanAccessibilitySpaceTime2023].

The objectives of this research are two-fold. First, we aim to theoretically bridge Mobility of Care with the 15-Minute City concept to re-imagine what local amenities matter and to define an associated accessibility and diversity of accessibility measures. Second, we take a data-driven approach to classify areas into 'caring 15-minute neighbourhoods' typologies and examine associated residential profiles in an empirical examination of the city of Hamilton, Canada.

@@

In the last decades, the need for more sustainable, healthier, and livable cities have received significant attention. Planners and decision-makers have proposed alternatives to mobility-based neighbourhood planning principles to ones that center proximity to urban functions [@pozoukidou15MinuteCityDecomposing2021]. In this context, the 15-Minute City is now in the public spotlight [@logan_xminute_2022; @moreno_introducing_2021]. The 15-Minute City was introduced in @moreno_introducing_2021 along with four dimensions: density (in terms of people per km$^2$), diversity (including mixed-land use and diversity of people), the temporal and spatial proximity to essential services, and digitalisation. However, the framework presented has been criticized within academic and planning circles [@guzmanProximityEnoughCritical2024]. Critiques highlight how the concept falls short in addressing the pre-existing structural forces and individual characteristics that drive inequalities, influencing who benefits or could benefit the most from such a paradigm [@dimarino15minuteCityConcept2023; @willberg15minuteCityAll2023]. A 15-Minute City for all people is an aspirational ideal, but how will existing mobility and accessibility inequalities be addressed. Without directed and context-specific solutions, this popular concept risks becoming an empty city branding exercise. [@pozoukidou15MinuteCityDecomposing2021; @gowerPlanningInnovationCity2022].

Reflecting the flexible and aspirational presentation of the 15-Minute City; cities that have adopted the concept have done so using a diverse range of definitions and tools along with indistinct universal approaches. A trail blazer was Portland, US in the Portland Plan [@city_of_portland_20minute_2010] of April 2012. Adopted before the '15-Minute City' concept of today, it aims to foster an inclusive city development based on prosperity, education, health, and equity over a 30 year horizon. The plan aims to promote neighbourhood self-sufficiency and connectivity to city centers and centres of employment. The progress report describes a high-level focus on equitable service delivery to all residents with equity concerning topics related to racial equity and people with disabilities [@portland_government_portland_2017], similarly taking on an "all populations" approach. Later, other cities adopted proximity-based goals using similar neutral approaches. For example, the 15-Minute City plan announced during the Paris mayor Anne Hidalgo’s 2020 re-election campaign, which helped popularize the concept in today's public discourse, emphasized six key social functions that should be easily accessible from any location [@ville_de_paris_paris_2022]. These locations included: housing, work, health care, groceries, education and leisure. The 15-Minute City concept inspired language in the agendas of other cities in the Western world such as Ottawa, in the Canadian context, who also adopted a 15-Minutes approach in their recent Official Plan [@ville_dottawa_quartier_2021]. @teixeiraClassifying15minuteCities2024's global review finds that the 15-Minute City concept is in early phases of implementation around the world and the diverse range of definitions, strategies, and instruments present a significant knowledge and implementation gap. In other words: the 15-Minute City is aspirational but how do we get there and who will benefit?

@@

The redefined bottom-up approach to community and neighbourhood planning has been adopted by New Urbanists, from which the 15-Minute City stems [@kissfazekasCircleParadigms15minute2022]; however, the effectiveness of these bottom-up processes remain to be seen due to the contemporary nature of the concept. Though, related research can be examined. In recent years, the question of _how can physical form be planned to enable an improved quality of life, for whom, and with what outcomes_ has occupied the urban planning research agenda. For instance, the examination of low-income households access to transportation and their likelihood of gaining employment [@blumenbergDriveWorkRelationship2017; @bastiaanssenDoesBetterJob2022] or the relationship between children's access to public transit and participation in after-school activities [@palmRolePublicTransit2020]. A new wave of researchers and practitioners focused on local and context-specific relationships with the proximity to destinations have also emerged [@silvaProximitycentredAccessibilityConceptual2023; @silvaRegionalAccessibilityUndermining2022]. As reviewed in the city plans that have adopted 15-Minute City approaches, the NUC's criticisms, and recent urban planning research, the question of _how to enable improved quality of life through urban built environment_ has shown to be highly context-sensitive, with the need for further investigation.

@@

In examining how to enable improved quality of life through urban built environment, accessibility measures have become an increasingly popular tool. These measures are a way to quantify the ease of reaching destinations from a given point in space and have been used to examine urban areas through just and sustainable city planning agendas [@vale2023accessibility; @handyAccessibilityIdeaWhose2020]. The 15-Minute City is an amenity-provision neighbourhood planning concept aimed at enabling the creation of urban environments that enhance life quality, making it well-suited for analysis using accessibility methodologies [@guzmanProximityEnoughCritical2024]. Recent works have applied accessibility measures to investigate the 15-Minute City across different geographic scopes. For instance, in Naples, Italy [@gaglioneUrbanAccessibility15minute2022], Barcelona, Spain [@graellsCityCitiesMeasuring2021], Vancouver, Canada [@hosford15minuteCityReach2022], and in urban areas across Europe [@vale2023accessibility]. The "cumulative opportunity" measure has been applied in many 15-Minute City examinations. This measure quantifies how many destinations can be reached from a point in space within a given travel time threshold, pairing well with examinations that have a normative travel time threshold, like the 15-Minute City. Furthermore, the cumulative opportunity measure is widely appreciated for its intuitive computation and popularity among practitioners [@handyAccessibilityIdeaWhose2020; @handyMeasuringAccessibilityExploration1997; @chengInvestigatingWalkingAccessibility2019]. However, accessibility measures other than the cumulative opportunity have also been applied, reflecting the diversity of measures in the literature [@guzmanProximityEnoughCritical2024].

Another method that compliments the assessment of the 15-Minute City regarding the diversity of opportunity types is entropy. The entropy measure, based on the Shannon-Wiener index of species diversity, expresses relative evenness within a sequence [@whittaker1972evolution]. In urban studies and planning literature, the flexible concept of entropy has been used to characterize land-use mix diversity [@frankLinkingObjectivelyMeasured2005; @ewingTravelBuiltEnvironment2010a], particularly in the context of non-work trips [@cerveroTravelDemand3Ds1997], walking as a mode [@luUrbanDensityDiversity2017; @mavoaIdentifyingAppropriateLanduse2018], and suburban sprawl [@randallGISBasedLand2015] as relevant to this work. It has also been used to understand mobility behaviour [@mcbride2020exploration; @montero2023applying; @montero2023role]. However, work incorporating entropy indices alongside accessibility analysis are not exceptionally common. There are examples of their use as parameters within accessibility scores for employment opportunities [@chengMeasuringUrbanJob2013; @daiIncorporatingJobDiversity2018] and, more recently, to describe the diversity in transit facilities [@yinIncorporatingFacilityDiversity2024]. Within the conceptualisation of the 15-Minute City, diversity of land-use is a key feature. In this way, the entropy measure theoretically complements the 15-Minute City.

Classification algorithms also show promise as a tool in the assessment of the 15-Minute City, as they have been useful in identifying mobility and spatial typologies within the transportation planning literature. Often, these algorithms take the form of machine learning approaches: for instance, @burkeGeospatialAnalysisFramework2022 uses spatially constrained multivariate clustering, to develop urban form typologies related to the 15-Minute City. The k-means algorithm is another popular method: @ganUnderstandingUrbanMobility2020 applies it to analyse travel behaviour and classify metro stations based on mobility patterns. Similarly, [@montero2023role; @montero2023applying] cluster travel behavior profile typologies representing different population segments and place of residence in the Barcelona Metropolitan Area. Combing classification techniques is also common: for instance, @delmelleDifferentiatingPathwaysNeighborhood2017 uses self-organizing maps (SOM) algorithm to group U.S. neighbourhoods by minimizing dissimilarity in attributes before applying k-means for further classification. Similarly, @victorianoTimeSpaceMoney2020 uses SOM to classify individuals' travel patterns according to the dissimilarity in mobility attributes and uses the Decision Tree algorithm to partition the data into interpretable classifications.

@@

Rather than focusing on _all_ destinations, it may valuable to examine those related to caring activities. This re-framing offers a feminist perspective on which urban functions matter, connecting well with the 15-Minute City concept. Caring activities, which fulfill the physical, psychological, and emotional needs of others, are among the most essential and fundamental activities in society [@ilo_care_2018]. Yet, they are the most unequal, undervalued, and even devalued activities worldwide. Conventionally, caring activities have been borne on women’s shoulders [@haydenGrandDomesticRevolution1982; @hochschild_second_2012]. According to @ilo_care_2018, women and girls perform more than three-quarters of the total amount of unpaid care worldwide, a gender gap that varies geographically [@RN5]. This unequal share of caring responsibilities leads to multifaceted gendered differences: in career development, profession selection, contract type, pay gap, and time poverty, as recognized by various international organisations [@EIGE2016; @ilo_care_2018]. In terms of spatial and transportation planning, almost one third of daily trips are for care purposes [@sanchezdemadariagaMobilityCareIntroducing2013; @sanchezdemadariagaMeasuringMobilitiesCare2019; @ravensbergen2023exploratory]. From this research motivation, Sánchez de Madariaga coined the term Mobility of Care in 2013 to refer to all travel required to sustain the needs of a household, such as grocery shopping, escorting children, travelling to health appointments, and running errands [@sanchezdemadariagaMobilityCareIntroducing2013]. While an undercurrent of research had examined these unique household-serving trips over the decades, her seminal work was the first to consider all these trips as one category and demonstrate how mobility of care is a significant proportion of daily travel.

The Mobility of Care concept also explicitly integrates inter-sectional equity considerations, a common criticism leveled at the 15-Minute City [@guzmanProximityEnoughCritical2024]. Perhaps unsurprisingly given the gendered division of care work discussed, women have consistently been found to complete more mobility of care trips than men. In one study, mobility of care comprised 32% of women’s daily trips compared to 28% of men’s. While this gap is significant, it was found to be far greater in lower income households where women complete 20% more care trips than men [@ravensbergen2023exploratory]. Sánchez de Madariaga not only shows how important these mobility of care trips are, but also highlights the ways in which “mobility of care is systematically under-represented in any analysis of urban transport” (p. 37). As a product of the masculinst bias in planning, transport surveys and tools often do not directly capture mobility of care, which re-enforces the idea that these trips are not a significant part of daily mobility. In this respect, the feminist perspective of the cities of proximity is still underestimated with only few examples on the topic [@gil_sola_choose_2022; @macintyre_her_2022].

@@

Hamilton also exhibits spatial disparities in social and economic indicators; their spatial distribution is visualised in @fig-Fig2. The densely populated inner city is characterised by lower average incomes, and a higher prevalence of households living under the low-income cutoff thresholds (LICO). Notably, the suburban areas of the city tend to have a greater proportion of children and a lower proportion of one parent households.
